---
title: "GeoAviR spatial models"
author: "Christian Roy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

##Startified models with distance.wrap

GeoAvir can be used to develop startified distance sampling model. 

The first step is to transform the survey lines form the project into a spatial object.

```{r,warning=FALSE,message=FALSE}
library(GeoAviR)
library(rgdal)

### Import and filter data
data(quebec)
d<-filterECSAS(quebec)
### Build a shapefile with transect starts
transect <- data.frame(lat=d$LatStart,lon=d$LongStart)
coordinates(transect) <- ~lon + lat
transect<-SpatialPointsDataFrame(transect,data=d[,"Count",drop=FALSE])
proj4string(transect)<-CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
```

The second step is to import a shapefile of the survey area and to associate each survey line to a zone in the survey area

```{r,warning=FALSE,message=FALSE}
### Import a spatial polygon
data(zonegulf)

### Overlay transects with gulf zones
zonegulf<-spTransform(zonegulf,CRS(proj4string(transect)))
x<-over(transect,zonegulf)
d$zone<-x$id
d$zone_area<-x$area
d<-d[!is.na(d$zone),]

### Build labels for samples.
d$SMP_LABEL<-paste(d$zone,d$Date,sep="_")
```

In our case, the observations made during each days are made of several WatchID (bouts). We will therefore use the day as the sample unit. Consequently, we need to to sum the WatchID within a day to calculate the effort for the day.

```{r,warning=FALSE,message=FALSE}
temp<-aggregate(WatchLenKm~SMP_LABEL,data=unique(d[,c("SMP_LABEL","WatchID","WatchLenKm")]),sum)
names(temp)[2]<-"SMP_EFFORT"
d<-merge(d,temp,sort=FALSE)
```

We then subset the dataset to keep only the variable of interest for the analysis and run the analysis. For this example we will use the observations of Black-legged Kittiwake (_Rissa tridactyla_). 

```{r,warning=FALSE,message=FALSE}
#Keep only the column of interest
d<-d[,c("zone","zone_area","Date","SMP_LABEL","SMP_EFFORT",
        "Distance","Count","Alpha","LatStart","LongStart")]
#eliminate duplicate lines for transect without observations
dd<-ddply(d,.(SMP_LABEL),function(i){sum(i$Count,na.rm=TRUE)})
#get the label name for transect without observations
dd<-dd[dd$V1==0,]
#keep only lines for empty transects or non-empty lines for non-empty transects
d<-d[(d$SMP_LABEL%in%dd$SMP_LABEL & !duplicated(d$SMP_LABEL)) |
       (!d$SMP_LABEL%in%dd$SMP_LABEL & !(d$Alpha=="")),]

d<-d[order(d$zone),]


###Run the analysis for the Gull species only
x <- distance.wrap(d,SMP_LABEL="SMP_LABEL",SMP_EFFORT="SMP_EFFORT",
                   DISTANCE="Distance",SIZE="Count",
                   units=list(Type="Line",Distance="Perp",Length_units="Kilometers",
                              Distance_units="Meters",Area_units="Square kilometers"),
                   breaks=c(0,50,100,200,300), estimator=list(c("HN","CO")),
                   detection="All",lsub=list(Alpha=c("BLKI")),split=TRUE,empty=NULL,
                   STR_LABEL="SMP_LABEL",STR_AREA="zone_area",stratum="STR_LABEL",
                   path="c:/temp/distance",
                   pathMCDS="C:/Distance 6",verbose=FALSE)

```

We can then proceed to look at the ouput of the analysis.

```{r}
x
summary(x)
```




