---
title: "R2MCDS Basic models"
author: "Christian Roy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

GeoAviR allows the to run some basic distance analysis on the dataset via the MCDS engine of Distance 6.2. The analyzes are currently restricted to line transect and to bin data. Distance sampling models are run via the `distance.wrap()` while strips transect models are analyzed with the function `strip.wrap()`. The distance model support 3 different key functions (Uniform, Half-normal and Hazard rate) and 3 different series adjustment (Cosine, Simple polynomial, Hermite polynomial). 

##Basic model with distance.wrap

If the user does not input any models via the estimator argument the function distance.wrap will fit by default the 6 possible detection function available in Distance 6.2:

- Uniform key function with Cosine adjustement
- Uniform key function with Hermine polynomial adjustement
- Half-normal key function with Cosine adjustement
- Half-normal key function with Hermine polynomial adjustement
- Hazard rate key function with Cosine adjustement
- Hazard rate key function with Simple polynomial adjustement

```{r,warning=FALSE,message=FALSE}
library(R2MCDS)
### Import and filter data
data(alcidae)
alcids <- mcds.filter(alcidae, transect.id = "WatchID", distance.field = "Distance", 
                          distance.labels = c("A", "B", "C", "D"), 
                          distance.midpoints = c(25, 75, 150, 250), effort.field = "WatchLenKm", 
                          lat.field = "LatStart", long.field = "LongStart", 
                          sp.field = "Alpha", date.field = "Date")

### Run analysis with the MCDS engine. Here, the WatchID is used as the sample.
x <- mcds.wrap(alcids, SMP_EFFORT="WatchLenKm",DISTANCE="Distance",SIZE="Count",
                   units=list(Type="Line",Distance="Perp",Length_units="Kilometers",
                              Distance_units="Meters",Area_units="Square kilometers"),
                   breaks=c(0,50,100,200,300), STR_LABEL="STR_LABEL", 
                   STR_AREA="STR_AREA",SMP_LABEL="WatchID", 
                   path="c:/temp/distance",
                   pathMCDS="C:/Distance 6",verbose=FALSE)
```
The distance.warp function will return a warning after fitting each model (not shown in the example above). These warning will include a "status" number that is associated with the status of the model by the <span style="font-family:Courier New"> MCDS </span> engine. The numbers should be intrepreted as:

- "1" the analysis ran without errors
- "2" the analysis ran with warnings
- "3" the analysis ran with errors
- "4" the analysis was not performed because of errors in the MCDS files
- ">4" a major error occured 

The function `distance.wrap()` will not return an output for any model which had a status above 2.A detailed description of each error message can be found in the Distance user guide (pages 333-344).  

Since we ran more than one model for a set of observations the model are stored into a `distanceList` object which consist of a list of `distanceFit` object. The output of each detection model can therefore be accessed by subsetting the list.
```{r,warning=FALSE,message=FALSE,fig.height=5,fig.width=5,fig.align='center'}
x
####summary of one model
summary(x[[2]])
predicted_hist(x[[2]])
```


The summary of a `distanceFit` object emulate the output of Distance 6.2 and present 3 tables. The first table contains all the information related to the encounter rate, the second table presents the parameters related to the detection probability, and the third table present the information related to the density/abundance in the survey area. A detailed description of the output can be found in the Distance user guide (Chapter 9 and page 330).


If the user is interested in keeping only the best model in the `distanceList` the function `keep.best.model()` will keep only the model with the lowest AICc value. If more than one model have the best AIC score the keep.best.model function will select one model randomly.

```{r,warning=FALSE,message=FALSE,fig.height=5,fig.width=5,fig.align='center'}
##### Keep the 'best' model in the list
x.best <- keep.best.model(x)
summary(x.best)
predicted_hist(x.best)
```

##Selecting the key function and adjustement with distance.wrap

It is possible for the user to select the key function and adjustment. For example, instead of running the 6 models by default the user may want to restrict his analysis to the half-normal key function with a cosine adjustment and the hazard rate key function with a polynomial adjustment. He can do so via the 'estimator' option. The estimator option consists of a list of elements where each element is a vector containing the first two letters of the key function and the first two letters of the expansion term. It is important to remember, however, that the list of detection key is restricted to the 6 key that have been presented above. 

```{r,warning=FALSE,message=FALSE}
library(R2MCDS)
### Import and filter data
data(alcidae)
alcids <- mcds.filter(alcidae, transect.id = "WatchID", distance.field = "Distance", 
                      distance.labels = c("A", "B", "C", "D"), 
                      distance.midpoints = c(25, 75, 150, 250), effort.field = "WatchLenKm", 
                      lat.field = "LatStart", long.field = "LongStart", 
                      sp.field = "Alpha", date.field = "Date")

### Run analysis with the MCDS engine. Here, the WatchID is used as the sample.
x <- mcds.wrap(alcids,SMP_EFFORT="WatchLenKm",DISTANCE="Distance",SIZE="Count",
               units=list(Type="Line",Distance="Perp",Length_units="Kilometers",
                            Distance_units="Meters",Area_units="Square kilometers"),
               breaks=c(0,50,100,200,300), SMP_LABEL="WatchID",
               STR_LABEL="STR_LABEL",STR_AREA="STR_AREA",
               estimator=list(c("HN","CO"),c("HA","PO")),
               path="c:/temp/distance",
               pathMCDS="C:/Distance 6",verbose=FALSE)
x
summary(x[[2]])
```

Once again it is possible to select the model with the lowest AIC value via the `keep.best.model()` function.

```{r,warning=FALSE,message=FALSE,fig.height=5,fig.width=5,fig.align='center'}
##### Keep the 'best' model in the list
x.best <- keep.best.model(x)
summary(x.best)
predicted_hist(x.best)
```
